Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-24 article.tex
checkpoint

% check dependency map => minimal d-map with distribution?
\textcolor{red}{This part is better, however there is still something
wrong with the theory.}
To a joint probability distribution $P_X$ corresponds a minimal dependency
graph. Each node of this graph is a random variable. Two random variables are
only connected if they are dependent. We consider that two neurons are
connected if  and only if,  there exists  in this dependency  graph an edge
that connects  directly the  two random  variables to  which they correspond.
This is equivalent to writing that two neurons $i$ and $j$ are  not  directly
connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
with  respect to all  other variables. We remind  that two random  variables
$X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
$X_j$ are  conditionally independent  with respect  to all  the other
variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
$X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no  information  about
$X_j$,  and vice-versa.

In the  particular case where $P_X$ is  a joint  Gaussian distribution
$\mathcal{N}(\mu; \Sigma)$ of mean  vector $\mu$ and covariance matrix
$\Sigma$, $X_i$  and $X_j$ are conditionally  independent with respect to the
other variables $X^{-i,j}$ if their \textit{partial correlation} statistic is
null~\citep{baba2004partial}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j |
X^{-i,j}} \neq 0$, where $\rho_{X_i, X_j | X^{-i,j}} \in [-1;1]$. Note that the
larger the value of this term, the stronger the dependence between variables
$X_i$ and $X_j$.

Our approach implicitly makes the additional assumption  that $P_X$ is  a
joint Gaussian distribution since it  outputs as  degree of association
between two neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$.
Moreover, this minimal dependency graph will be close to G undirected
especially as this assumption is true. Note also that the partial correlation
statistic is symmetric  (i.e., $\rho_{X_i, X_j | X^{-i,j}} = \rho_{X_j, X_i |
X^{-j,i}}$) and therefore does not account for the direction of the
interaction without impacting the performance. The interest to not consider directivity will be discussed in section~\ref{sec:results}.

We detail now our procedure for estimating $\rho_{X_i, X_j | X^{-i,j}}$ from
the data. \textcolor{red}{DE finds that this sections is too complicated}
The partial correlation statistics is computed through the inverse
$\Sigma^{-1}$ of the covariance matrix $\Sigma$, also known as the
concentration or precision matrix. Computationally, they can be estimated
efficiently in $O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal
components extracted through PCA~\citep{bishop2006pattern} on the processed
data. In our experiments, $M$ is set to $0.8 p$, thereby allowing us to further
filter the the data by removing the contribution of principal components likely
due to noise.

To highlight the benefits of measuring the partial dependence conditional to
all other variables, we compare in Section~\ref{sec:results} with the
\textit{Pearson correlation} statistics, which measure instead the linear
dependence between variables, independently of the others. Accordingly, $X_i$
and $X_j$ are independent if their Pearson correlation statistic is zero, that
is if
\begin{equation}
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
\end{equation}

>>>>

To a joint probability distribution $P_X$ corresponds a
graph where (i) each node is a random variable and (ii) two nodes are
only connected if they correspond to two dependent variables. We consider that two neurons are
connected if  and only if,  there exists  in this dependency  graph an edge
that connects  directly the  two random  variables to  which they correspond.
In this context, it is equivalent to writing that two neurons $i$ and $j$ are  not  directly
connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
with  respect to all  other variables. We remind  that two random  variables
$X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
$X_j$ are  conditionally independent  with respect  to all  the other
variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
$X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no information  about
$X_j$,  and vice-versa.

In the  particular case where $P_X$ is  a joint  Gaussian distribution
$\mathcal{N}(\mu; \Sigma)$ of mean  vector $\mu$ and covariance matrix
$\Sigma$, $X_i$  and $X_j$ are conditionally  independent with respect to the
other variables $X^{-i,j}$ if their \textit{partial correlation} statistic is
null~\citep{baba2004partial}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0 \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$ is also known as the precision matrix or concentration. 
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j |
X^{-i,j}} \neq 0$, where $\rho_{X_i, X_j | X^{-i,j}} \in [-1;1]$. Note that the
larger the value of this term, the stronger the dependence between variables
$X_i$ and $X_j$.

Our approach implicitly makes the additional assumption  that $P_X$ is  a
joint Gaussian distribution since it  outputs as degree of association
between two neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$.
Note also that the partial correlation
statistic is symmetric  (i.e., $\rho_{X_i, X_j | X^{-i,j}} = \rho_{X_j, X_i |
X^{-j,i}}$). Therefore, our approach cannot identify the direction of the interaction. We will see why this only affects slightly its performances, with respect to the metric used in the Connectomics challenge, in Section~\ref{sec:results}.

We detail now our procedure for estimating partial correlation statistics ($\rho_{X_i, X_j | X^{-i,j}}$) from
the data. Besides the straightforward computation, a natural procedure for estimating this precision matrix is first computing an empirical estimation of the covariance matrix and then taking the inverse, applying equation \ref{eq:inverse}. 
Yet, an approximation can be obtained by only considering the first $M$principal components. Notice that we could have expected that a approximation of the inverse would have degraded the performance. However, the opposite was observed. We conjecture that by only taking into account the $M$ principal components filter out noise. In our experiments, $M$ is set to $0.8p$ (see discussion in Appendix~\ref{app:REF} \textcolor{red}{REF}).

% Computationally, they can be estimated
% % efficiently in $O(M^3)$ (instead of $O(p^3)$) 
% from the $M$ first principal
% components extracted through PCA~\citep{bishop2006pattern} on the processed
% data. While speeding up computation, it allows to further 
% filter the the data by removing the contribution of components that are respectively
% due to redundancy or noise, i.e., with an eigen value respectively is equal to zero or close to zero.

To highlight the benefits of measuring the partial dependence conditional to
all other variables, we compare in Section~\ref{sec:results} with the
\textit{Pearson correlation} statistics, which measure instead the linear
dependence between variables, independently of the others. Accordingly, $X_i$
and $X_j$ are independent, without any conditioning, if their Pearson correlation statistic is zero, that
is if
\begin{equation}
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
\end{equation}
